Comparing methods for estimating parasite-induced host mortality from distributional data: implementation and limitations

### 2015-04-27 docs/parasite_mortatlity_manuscript.tex
Made edits before sending to Cherie

In this study, we have
three primary goals. First we wish to test reliability of the Adjei Method for
answering the aforementioned questions regarding PIHM.  Second, we propose a
novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods ability to detect PIHM on empirical data.  Third, we
discuss the limitations of inferring PIHM from distributional data alone and
whether any method for inferring PIHM has a place in the future of
parasitology.

@@

To estimate this function the Adjei Method proceeds as follows.  First, it
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_0^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$. Therefore equation \ref{eq:concept} can be written as

@@

To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.

@@

The ability of the Adjei Method to detect PIHM given that it was occurring in a
system (i.e. the power) was close to 1 for larger sample sizes and tended to
decrease as sample size decreased (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  The Likelihood Method had a power close to
one for all parameter combinations and sample sizes considered.  With gradual
survival functions (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2), the Likelihood Method showed a
decreased power to detect PIHM for small sample sizes.

@@

In terms of the host survival function, the Likelihood Method gave
asymptotically unbiased estimates of $a$ and $b$ as sample size increased for
all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
Likelihood Method tended to produce severely biased estimates of $a$ and $b$.
This was generally more pronounced for steeper survival functions and more
aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
biased estimates of $a$ and $b$ across all sample sizes, with the bias
consistently being larger when the survival function was steeper. The bias of
the Adjei Method's estimate of $a$ also increased as $\mu_p$ increased (Figure \ref{fig:biasa}, SI2 Fig 5, 6).

@@

To attempt to ameliorate the flaws in the Adjei Method, we proposed a more
general method to determine both whether or not PIHM is occurring in a system
and to quantify the survival function.  We show that this method is
asymptotically unbiased when estimating the host-survival function for all of
the parameter space that we explored and we found that it produces unbiased and
precise estimates of the $LD_{50}$ for small, realistic sample sizes. Moreover,
this novel method has a Type I error rate close to the pre-set level of $\alpha = 0.05$ and high power for detecting PIHM for realistic samples sizes.
  However, we note that
the Likelihood Method produces seriously biased estimates of the host survival
function ($a$ and $b$) for sample sizes typically observed in many host-
parasite studies.  The bias was most severe for steep host survival functions,
due to large changes in the values of $a$ and $b$ only slightly changing an
already steep survival function.  Given these results, neither the
Likelihood Method or the Adjei Method could confidently recover the exact shape of the host survival function for small, realistic sample sizes.

We also fit both the Likelihood Method and the Adjei Method to empirical data to
determine whether they could detect PIHM that had been previously reported
based on visual assessments.  Consistent with our simulation results, we found
that the Adjei Method tended to detect PIHM where it had not been previously
reported, while the Likelihood Method's detection of PIHM was consistent with
previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is fundamentally
flawed.  We recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.

While we have improved upon the previously existing methods for answering questions
about PIHM, we cannot belie the fact that estimating PIHM from observational
data alone is ladened with assumptions and difficulties \citep{McCallum2000a}.
The most fundamental assumption of all methods for estimating PIHM is that the
shape of the pre-mortality host-parasite distribution is known and follows a negative binomial distribution.
While there is substantial empirical and theoretical evidence to justify the
use of the negative binomial distribution as the pre-mortality distribution for
macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
widely recognized that different processes can lead to a variety of
distributions of parasites across hosts \citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the
critical assumption of the pre-mortality distribution is not that the processes
leading to the pre-mortality distribution generate a negative binomial
distribution, but rather that the pre-mortality distribution is well-fit by a
negative binomial. The extreme flexibility of the negative binomial
distribution makes it a reasonable candidate distribution for the pre-mortality
distributions.  Therefore, we do not see this assumption as central problem in
any of the proposed methods.

However, to use the pre-mortality distribution to infer whether or not the PIHM
is occurring in a system requires an explicit assumption about the host
survival function and the shape of the post-mortality distribution.  Regarding
the host-survival function, all currently proposed methods of PIHM assume that
the host-survival function is such that uninfected individuals and individuals
with low parasite intensity experience essentially no PIHM.
\cite{Lanciani1989} illustrated the importance of this assumption by showing
that when hosts experienced a linear decrease in survival probability the
Crofton Method could not detect PIHM.   As the most fundamental models of host-
parasite dynamics assume a linear decrease in host survival probability with
increasing parasite intensity \citep{AndersonandMay1978}, the failure of these methods
to detect this relationship is a significant disconnect between empirical and
theoretical disease ecology [though I haven't explicitly tested this with the likelihood method].  However, empirical work has shown that non-linear
functions of host survival are not uncommon in host-parasite systems
\citep{Benesh2011}, so this assumption alone does not preclude the use of PIHM
methods on empirical data.

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each methods $a$ estimate over 150 simulations. The second column gives the precision of each methods $a$ estimate over 150 simulations.}

>>>>

In this study, we have
three primary goals. First we wish to test reliability of the Adjei Method for
answering the aforementioned questions regarding PIHM.  Second, we propose a
novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods' ability to detect PIHM on empirical data.  Third, we
discuss the limitations of inferring PIHM from distributional data alone and
whether any method for inferring PIHM has a place in the future of
parasitology.

@@

To estimate this function the Adjei Method proceeds as follows.  First, it
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$. Therefore equation \ref{eq:concept} can be written as

@@

To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

@@

The ability of the Adjei Method to detect PIHM given that it was occurring in a
system (i.e. the power) was close to 1 for larger sample sizes and tended to
decrease as sample size decreased (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  The Likelihood Method had a power close to
one for all parameter combinations and sample sizes considered.  With gradual
survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2).

@@

In terms of the host survival function, the Likelihood Method gave
asymptotically unbiased estimates of $a$ as sample size increased for
all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
Likelihood Method tended to produce severely biased estimates of $a$.
This was generally more pronounced for steeper survival functions and more
aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
biased estimates of $a$ across all sample sizes, with the bias
consistently being larger when the survival function was steeper. The bias of
the Adjei Method's estimate of $a$ also increased as $\mu_p$ increased (Figure \ref{fig:biasa}, SI2 Fig 5, 6).

@@

To attempt to ameliorate the flaws in the Adjei Method, we proposed a more
general method to determine both whether or not PIHM is occurring in a system
and to quantify the survival function.  We show that this method is
asymptotically unbiased when estimating the host-survival function for all of
the parameter space that we explored and we found that it produces unbiased and
precise estimates of the $LD_{50}$ for small, realistic sample sizes. Moreover,
this novel method has a Type I error rate close to the pre-set level of $\alpha = 0.05$ and high power for detecting PIHM for realistic samples sizes.
  However, we note that
the Likelihood Method produces seriously biased estimates of the host survival
function ($a$ and $b$) for sample sizes typically observed in many host-parasite studies.  The bias was most severe for steep host survival functions,
due to large changes in the values of $a$ and $b$ only slightly changing an
already steep survival function.  Given these results, neither the
Likelihood Method nor the Adjei Method could confidently recover the exact shape of the host survival function for small, realistic sample sizes.

We also fit both the Likelihood Method and the Adjei Method to empirical data to
determine whether they could detect PIHM that had been previously reported
based on visual assessments.  Consistent with our simulation results, we found
that the Adjei Method tended to detect PIHM where it had not been previously
reported, while the Likelihood Method's detection of PIHM was consistent with
previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is fundamentally
flawed and we recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.

While we have improved upon the previously existing methods for answering questions
about PIHM, we cannot belie the fact that estimating PIHM from observational
data alone is ladened with assumptions and difficulties \citep{McCallum2000a}.
The most fundamental assumption of all methods for estimating PIHM is that the
shape of the pre-mortality host-parasite distribution is known and follows a negative binomial distribution.
While there is substantial empirical and theoretical evidence to justify the
use of the negative binomial distribution as the pre-mortality distribution for
macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
widely recognized that different processes can lead to a variety of
distributions of parasites across hosts \citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the
critical assumption of the pre-mortality distribution is not that the processes
leading to the pre-mortality distribution generate a negative binomial
distribution, but rather that the pre-mortality distribution is well-fit by a
negative binomial distribution. The extreme flexibility of the negative binomial
distribution makes it a reasonable candidate distribution for the pre-mortality
distributions.  Therefore, we do not see this assumption as central problem in
any of the proposed methods.

However, to use the pre-mortality distribution to infer whether or not the PIHM
is occurring in a system requires an explicit assumption about the host
survival function and the shape of the post-mortality distribution.  Regarding
the host-survival function, all currently proposed methods of PIHM assume that
the host-survival function is such that uninfected individuals and individuals
with low parasite intensity experience essentially no PIHM.
\cite{Lanciani1989} illustrated the importance of this assumption by showing
that when hosts experienced a linear decrease in survival probability the
Crofton Method could not detect PIHM.   As the most fundamental models of host-parasite dynamics assume a linear decrease in host survival probability with
increasing parasite intensity \citep{AndersonandMay1978}, the failure of these methods
to detect this relationship is a significant disconnect between empirical and
theoretical disease ecology.  However, empirical work has shown that non-linear
functions of host survival are not uncommon in host-parasite systems
\citep{Benesh2011}, so this assumption alone does not preclude the use of PIHM
methods on empirical data.

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each method's $a$ estimate over 150 simulations. The second column gives the precision of each method's $a$ estimate over 150 simulations.}

### 2015-04-27 docs/supplementary_material.tex
Made edits before sending to Cherie

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each methods $a$ estimate over 150 simulations. The second column gives the precision of each methods $a$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each methods $a$ estimate over 150 simulations. The second column gives the precision of each methods $a$ estimate over 150 simulations.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
    \caption{A comparison of this papers implementation of the Crofton Method with the results given in \cite{Crofton1971a}.  Three of the 6 stations fit by in \citep{Crofton1971a} are shown here and all three show that the two implementations give very similar results.}
    \label{fig:crof_test}

>>>>

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each method's $a$ estimate over 150 simulations. The second column gives the precision of each method's $a$ estimate over 150 simulations.}

@@

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each method's $a$ estimate over 150 simulations. The second column gives the precision of each method's $a$ estimate over 150 simulations.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
    \caption{A comparison of this papers implementation of the Crofton Method with the results given in \cite{Crofton1971a}.  Three of the 6 stations fit by \cite{Crofton1971a} are shown here and all three show that the two implementations give very similar results.}
    \label{fig:crof_test}
